---
title: "Putting Google Fonts in R graphs"
author: "Cecina Babich Morrow"
date: 2019-10-11T21:13:14-05:00
categories: ["R", "Google"]
tags: ["R", "Google"]
draft: false
subtitle: How to use a Google Font in R.
summary: How to use a Google Font in R.
image:
  placement: 1
  caption: "In keeping with my Lego theme, this statue from Nathan Sawaya's exhibition at the New York Hall of Science represents some of the frustrations of adding fonts to R."
  focal_point: ''
  preview_only: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Inspiration for this post

I've been working on creating a visualization cookbook for R graphics (check out [this great example](https://bbc.github.io/rcookbook/) from the BBC to see what I'm eventually going for). As part of this process, I wanted to be able to change the font of my plots to [Source Sans Pro](https://fonts.google.com/specimen/Source+Sans+Pro), which is a Google font.

## `sysfonts` package

The first step here was to install and load the `sysfonts` package:

```{r}
library(sysfonts)
```

### XQuartz

If you end up getting the following error when loading `sysfonts`: `Reason: image not found`, you might need to do some additional work. This happened to me initially, and after doing some Googling, I found [this GitHub issue](https://github.com/yixuan/showtext/issues/13). Installing XQuartz was recommended as a fix, particularly for Macs, so I downloaded it from [here](https://www.xquartz.org/). After downloading XQuartz, I uninstalled, reinstalled, and loaded `sysfonts` and things went smoothly.

## Downloading font to your computer

As I learned from this [super helpful GitHub repo](https://github.com/kjhealy/sourcesans/blob/150207b18f4fbcc4c954715f269f8b20ce834a54/R/sourcesans.r), you need to install the font on your system as well as within R.

### Mac

I went through this process on a Mac, so this is what I ended up doing to install the font. I went to the [Google Fonts page for the font](https://fonts.google.com/specimen/Source+Sans+Pro) and clicked on "Select this font". Then after clicking on "1 Family Selected" at the bottom, I downloaded the font by clicking on the download icon. This downloaded a zip drive into my Downloads folder. To install it as a font, I extracted the zip folder (by opening the zip drive). In a different Finder window, I opened my Applications folder and found Font Book. Then I dragged the zip folder over into Font Book.

### PC

I modified the above instructions from these instructions on [Flourish Online](https://flourishonline.com.au/download-google-fonts-mac-pc/), which also has PC instructions. If anyone gives these a try and they don't work, let me know and I'll try to do some troubleshooting!

## `font_add_google`

Next, I used the `font_add_google` function from `sysfonts` to download the Source Sans Pro fonts:

```{r}
font_add_google("Source Sans Pro")
```

If you want to install a different Google Font, you can run `font_families_google()` to see the list of family names of fonts currently available in Google Fonts:

```{r}
head(font_families_google())
```

## Example

To continue with my Lego theme, I'll demonstrate using this font on data from the `legocolors` package. We can make the following graph of approximate lego brick availability based on the year a brick was released (colored by the brick color!):

> `r emo::ji("exclamation_mark")` **Versions >= 0.3.0 of `legocolors` no longer include `bl_bp`**: This code requires an installation of an earlier version of `legocolors`, which can be accomplished using `devtools::install_version("legocolors", version = "< 0.3")`


```{r message = FALSE}
library(legocolors)
library(ggplot2)

ggplot(legocolors, aes(x = year_released, y = bl_bp, colour = hex)) +
  geom_point() +
  scale_color_manual(values = legocolors$hex) +
  theme_minimal() +
  theme(legend.position = "none", text = element_text(family = "Source Sans Pro"), plot.title = element_text(face = "bold")) +
  labs(x = "Year released", y = "Brick availability", title = "Lego availability", subtitle = "All of the text is now in Source Sans Pro!")
```

## Troubleshooting

This process is a little bit finicky, so there are certain errors that crop up frequently. For me, just restarting RStudio and/or my computer (or updating the computer, if necessary) has solved these. The most common errors are `No font could be found for family "Source Sans Pro"` and `"Error in grid.Call(C_textBounds, as.graphicsAnnot(x$label), x$x, x$y, : polygon edge not found"`, both of which relate to R not being able to find the font. If the font is indeed installed on your computer (you can check this in Font Book, for a Mac), then restarting things should help R find the font.


